---
title: "Seasonal and Event Behaviour"
format: html
params:
  lake_name: "Lake Rotorua"
  lat: -38.114908
  lon: 176.316662
  buffer_km: 10
  lake_id: "rotorua"
  reference: "Airport_1770"
  targets: ["ERA5", "Buoy", "Town_40177", "VCS_On"]
  vars: ["Temp_C", "Precip_mm", "Wind_Spd_ms", "RadSWD_Wm2"]
  wet_threshold_mm: 1
  precip_event_threshold: 0.1
  windy_top_pct: 0.10
  windy_threshold_ms: 10
  window_days: 30
---

This page focuses on behaviour: when datasets agree on events, where they miss each other (zeros vs non-zeros, timing offsets, extremes), and how seasonality shows up.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

```{r}
#| echo: false
library(glue)
library(htmltools)
library(gt)
library(plotly)
library(here)
library(tidyverse)
library(lubridate)
library(slider)
library(reactable)

# Source shared helpers
source(here("scripts", "01_metrics_helpers.R"))
source(here("scripts", "03_analysis_helpers.R"))
source(here("scripts", "04_analysis_plotting.R"))
```

```{r}
#| echo: false
# Data setup (mirrors rotorua.qmd)
file_map <- list(
  ERA5         = here("data/processed/rotorua_era5_daily.csv"),
  Buoy         = here("data/processed/rotorua_buoy_daily.csv"),
  Town_40177   = here("data/processed/rotorua_town_40177_daily.csv"),
  VCS_On       = here("data/processed/rotorua_vcs_on_daily.csv"),
  Airport_1770 = here("data/processed/rotorua_airport_1770_daily.csv")
)

read_from_map <- function(name) {
  readr::read_csv(file_map[[name]], show_col_types = FALSE) |>
    mutate(Date = as.Date(Date))
}

ref_name <- params$reference
target_names <- params$targets

ref_df <- read_from_map(ref_name)
targets_list <- setNames(lapply(target_names, read_from_map), target_names)

wet_threshold_mm <- params$wet_threshold_mm
precip_event_threshold <- params$precip_event_threshold
windy_top_pct <- params$windy_top_pct
windy_threshold_ms <- params$windy_threshold_ms
window_days <- params$window_days

# only keep variables present in all datasets to prevent render errors
vars_requested <- params$vars
vars_available <- Reduce(
  intersect,
  c(list(names(ref_df)), lapply(targets_list, names))
)
vars <- intersect(vars_requested, vars_available)
```

## Precipitation event structure

These plots show how precipitation volumes and timing align across datasets.

```{r}
plot_precip_hydrograph(ref_df, targets_list, show_7day_sum = TRUE)
```

```{r}
plot_precip_cdf(ref_df, targets_list)
```

```{r}
plot_precip_wetday_distribution(ref_df, targets_list, threshold_mm = wet_threshold_mm)
```

## Event detection and mismatches

Use these to spot zero vs non‑zero disagreements, misses/false alarms, and timing offsets during wet events. POD: Probability detection FAR: False alarm ratio CSI: Critical success inder Bias score:

```{r}
#| echo: false
skill_precip <- event_skill_all_targets(
  ref_df, targets_list,
  var = "Precip_mm",
  threshold = precip_event_threshold,
  direction = ">="
)

reactable(
  skill_precip,
  sortable = TRUE,
  striped = TRUE,
  pagination = FALSE,
  highlight = TRUE,
  defaultPageSize = 10
)
```

```{r}
plot_event_outcomes(
  skill_precip,
  title = paste0("Precip events (\u2265 ", precip_event_threshold, " mm/day): outcomes vs reference")
)
```

```{r}
plot_event_skill_scores(
  skill_precip,
  title = paste0("Precip events (\u2265 ", precip_event_threshold, " mm/day): skill vs reference")
)
```

```{r}
plot_false_alarm_magnitude(ref_df, targets_list, threshold_mm = precip_event_threshold)
```

## Wind and precipitation timing drifts

Rolling diagnostics reveal gradual offsets or drift through time (look for breaks away from 0 bias or changing correlation). This answeres "does it shift seasonally?"

```{r}
purrr::iwalk(targets_list, ~ print(plot_rolling_diagnostics(ref_df, .x, .y, "Wind_Spd_ms", window_days = window_days)))
```

```{r}
purrr::iwalk(targets_list, ~ print(plot_rolling_diagnostics(ref_df, .x, .y, "Precip_mm", window_days = window_days)))
```

## Seasonal behaviour

Climatology and seasonal bias plots highlight over/under-capture of extremes across the year.

```{r}
plot_monthly_climatology(ref_df, targets_list, "Temp_C")
```

```{r}
plot_monthly_climatology(ref_df, targets_list, "Wind_Spd_ms")
```

```{r}
plot_monthly_climatology(ref_df, targets_list, "Precip_mm")
```

```{r}
season_defs <- season_defs_default
plot_seasonal_bias(ref_df, targets_list, "Temp_C", season_defs = season_defs)
```

```{r}
plot_seasonal_scatter(ref_df, targets_list, "Wind_Spd_ms", season_defs = season_defs)
```

```{r}
plot_seasonal_scatter(ref_df, targets_list, "Precip_mm", season_defs = season_defs)
```